---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Age-structured population models'
    slide-number: true
lang: en
---



# Introduction {visibility="hidden"}

## {content-align="center" style="font-size: 160%;"}


<center>

**Age-structured and stage-structured <br> population models**
    
![](figs/AMRE-HY-male){height="2.7cm"}
![](figs/AMRE-SY-male){height="2.7cm"}
![](figs/AMRE-ASY-M){height="2.7cm"}

</center>




# Introduction


## Motivation
  
  
  
  % * These facts have important management implications. <br>
%    * Harvesting larger numbers of the age group that most
%      positively influences population growth = bad idea
%    * If habitat degradation affects an influential age group, we may be putting the
%      population at risk
%    * Will be more successful if we re-establish the most influential age group
%    * Population control - May want to target the most influential age group for
%      elimination
%    * Take Home - Determining the most influential age is of paramount importance for
%      knowing what management actions should be conducted!
%  




## Human mortality risk
  
  [width=0.8\textwidth]{figs/Mortality_by_age} <br>




## Human age distributions

{ \bf Declining populations have relatively more old individuals than
  growing populations <br>}

<center>
    [width=\textwidth]{figs/AgeStructures}
</center>




## Three age class example

Abundance of age class $i$ in year $t$ is denoted by $n_{i,t}$.
  
The model for $n_{i,t+1}$ depends on: 
  
- Age class survival $s_{i}$, and
- Age class fecundity $f_{i}$
- Fecundity is the birth rate ($b_i$) multiplied by survival
    
- In pre-reproductive surveys, $f_i = b_i \times s_0$
- In post-reproductive surveys, $f_i = b_i \times s_i$
    

## How does this population grow?
  
  <center>
    \begin{tabular}{cl}
      \hline
      Age class & Equation <br>
      \hline
      1 & $n_{1,t+1} = n_{1,t} \times f_1 + n_{2,t} \times f_2 + n_{3,t} \times f_3$ <br> 
      2 & $n_{2,t+1} = n_{1,t} \times s_{1}$ <br> 
      3 & $n_{3,t+1} = n_{2,t} \times s_{2}$ <br>
      \hline
    \end{tabular}
  </center>
  
  
  
According to this particular age-structured model, all animals die
after spending one time interval in the third age class. <br>  
  
  
In stage-structured models, animals can stay in a class for
more than one time interval. 


## Age-class abundance and age distribution
  
* <1-> Suppose initial abundance in the 3 age classes is:
       
        * $n_{1,1} = 50$ (Age class 1, juveniles)
        * $n_{2,1} = 40$ (Age class 2, subadults)
        * $n_{3,1} = 10$ (Age class 3, adults)
        * $N_1$ = 100
      
    * []
    * <2-> This implies an initial age distribution of:
       
%        * $c_{1,0} = n_{1,0}/\sum_{i=1}^3 n_{i,0} = 0.5$
%        * $c_{2,0} = n_{2,0}/\sum_{i=1}^3 n_{i,0} = 0.4$
%        * $c_{3,0} = n_{3,0}/\sum_{i=1}^3 n_{i,0} = 0.1$
        * $c_{1,1} = n_{1,1}/N_1 = 0.5$
        * $c_{2,1} = n_{2,1}/N_1 = 0.4$
        * $c_{3,1} = n_{3,1}/N_1 = 0.1$
      
  
  
**An age distribution describes the proportion of individuals in each age class**



## Age distribution

```{r}
#| label: struc1
#| include: true
#| echo: false
#| fig.width: 6
#| fig.height: 6
#| out.width: '0.85<br>textwidth'
#| fig.align: 'center'
barplot(c(0.5, 0.4, 0.1), cex.lab=1.3, cex.names=1.3, cex.main=1.3,
        names=c("Juveniles", "Subadults", "Adults"),
        ylab="Proportion in each age class",
        main="Initial age distribution")
```




  ## Example

<center> 
    Let's choose values of $s_i$ and $f_i$
</center>
  
  \begin{tabular}{cccc}
    \hline
    Age class & Initial population size & Survival probability & Fecundity rate <br>
              & $n_{i,1}$  & $s_i$ & $f_i$ <br>
    \hline
    1 & 50 & 0.5 & 0.0  <br>
    2 & 40 & 0.6 & 0.8  <br>
    3 & 10 & 0.0 & 1.7  <br>
    \hline
  \end{tabular}




## Population size, $n_{i,t$}

```{r}
#| label: proj1
#| include: false
#| echo: false
#| results: 'hide'
T <- 30
N <- matrix(NA, 3, T+1)
N[,1] <- c(50, 40, 10)
phi <- c(0.5, 0.6, 0.0)
gamma <- c(0, 0.8, 1.7)
A <- matrix(c(gamma, phi[1], 0, 0, 0, phi[2], 0), 3, 3, byrow=TRUE)
for(t in 1:T) {
    N[,t+1] <- A %*% N[,t]
}
TT <- matrix(0:T, 1, T+1)
matplot(t(TT), t(N), type="o", pch=16, xlab="Time", ylab="Population size",
        cex.lab=1.3, ylim=c(0, 60), col=c("black", "orange", "purple"))
legend(20, 60, c("Age class 1", "Age class 2", "Age class 3"),
       col=c("black", "orange", "purple"), pch=16, lty=1:3)
```

<center>
  [width=0.85\textwidth]{figure/proj1-1}
</center>






## Growth rates, $\lambda_{i,t=n_{i,t}/n_{i,t-1}$}

```{r}
#| label: lambda1
#| include: false
#| echo: false
lambda <- N[,2:(T+1)] / N[,1:T]
matplot(t(lambda), type="o", pch=16,
        xlab="Time", ylab="Population growth rate (lambda)",
        cex.lab=1.3,
        col=c("black", "orange", "purple"))
legend(20, 2.4, c("Age class 1", "Age class 2", "Age class 3"),
       col=c("black", "orange", "purple"), pch=16, lty=1:3)
```

<center>
  [width=0.85\textwidth]{figure/lambda1-1}
</center>





## Age distribution, $c_{i,t=n_{i,t}/N_t$}

```{r}
#| label: prop1
#| include: false
#| echo: false
Nt <- colSums(N)
C <- sweep(N, 2, Nt, "/")
matplot(t(TT), t(C), type="o", pch=16, xlab="Time",
        cex.lab=1.3,
        ylab="Proportion in age class",
        col=c("black", "orange", "purple"), ylim=c(0, 1))
legend(20, 1, c("Age class 1", "Age class 2", "Age class 3"),
       col=c("black", "orange", "purple"), pch=16, lty=1:3)
```

<center>
  [width=0.85\textwidth]{figure/prop1-1}
</center>



## Important things to note
  
%  * Age distribution converges to a {\bf stable age
    distribution} when survival and fecundity rates are constant. <br>
  
  
  % * Stable age distribution is the proportion of individuals in each age
  class when the population converges. <br>
  
  
  % * Growth rates of each age class differ at first, but converge
  once the stable age distribution is reached. <br>
  
  
  % * Asymptotic growth rate is $\lambda$ (without subscript). <br>
  
  
  % * Growth rate at the stable age distribution is the same for
  all age classes, and it is geometric! <br>
%    * But abundance continues to change
%  


# Matrix Models


## Matrix models
  
%  [<+->]
%  * Matrix models aren't actually ``new'' models. <br>
  
  
  % * They are the same models we have been talking about. <br>
  
  
  % * But, they make it much easier to compute important
  quantities like $\lambda$ and {\it reproductive value}. <br>
%    * Everyone will think you are cool if you know what a Leslie
%      matrix is
%  





## Matrix models vs life tables

::: columns
%    
    \begin{column}[T]{0.5\textwidth}
      {\bf Matrix models}
      
        * <1-> Age is discrete
        * <2-> Age class is denoted by $i$
%        * <3-> Individuals either advance to the next age class or die
%        * <3-> An age class is not necessarily a 1 year time period
        * <3-> Each age class can have its own vital rates
%        * We usually only care about females
      
    :::
    \begin{column}[T]{0.5\textwidth}
      {\bf Life tables}
      
        * <1-> Age is continuous
        * <2-> Actual age is denoted by $x$
        * <3-> Each age can have its own vital rates
      
    :::
  :::
  <center>
    [width=0.9\textwidth]{figs/age-class-diagram}
  </center>





## What is a matrix?

{\bf Definition}: A matrix is a rectangular array of numbers
  
    * Usually denoted by an uppercase, bold letter
    * Either square or rounded brackets are used
    * Usually, rows are indexed by $i$, columns by $j$
  
  

  
  Example of a $3 \times 4$ matrix: \par
  <center>
    
    \[
    { \bf A} =
    \begin{bmatrix}
      a_{1,1} & a_{1,2} & a_{1,3} & a_{1,4} <br>
      a_{2,1} & a_{2,2} & a_{2,3} & a_{2,4} <br>
      a_{3,1} & a_{3,2} & a_{3,3} & a_{3,4}
    \end{bmatrix}
    \]
  </center>




## Leslie matrix
  
{\bf What is it?}
  
    * Square matrix %$s \times s$
    * Fecundities on first row
    * Survival probs on lower off-diagonal
  
  
Example:
  
<center>
    \[
    {\bf A} =
    \begin{bmatrix}
      f_1 & f_2 & f_3 & f_4 <br>
      s_1 & 0 & 0 & 0 <br>
      0 & s_2 & 0 & 0 <br>
      0 & 0 & s_3 & 0
    \end{bmatrix}
    \]
</center>



## How do we use a Leslie matrix?
  
These two expressions are equivalent: \par
  
  \begin{tabular}{cl}
    \hline
    Age class & Equation <br>
    \hline
    1 & $n_{1,t+1} = n_{1,t} \times f_{1} + n_{2,t} \times f_{2} + n_{3,t} \times f_3$ <br>
    2 & $n_{2,t+1} = n_{1,t} \times s_{1}$ <br>
    3 & $n_{3,t+1} = n_{2,t} \times s_{2}$ <br>
    \hline
  \end{tabular}
  
  { AND \par}
  
  \[
    {\bf n}_{t+1} = {\bf A}\times {\bf n}_{t}
  \]





  ## Matrix multiplication
  
  <center>
    \[
    \uncover<2>{ %
    \begin{bmatrix}
      aw + bx + cy + dz <br>
      ew + fx + gy + hz <br>
      iw + jx + ky + lz <br>
      mw + nx + oy + pz
    \end{bmatrix}
    } =
    \begin{bmatrix}
      a & b & c & d <br>
      e & f & g & h <br>
      i & j & k & l <br>
      m & n & o & p
    \end{bmatrix}
    \times
    \begin{bmatrix}
      w <br>
      x <br>
      y <br>
      z
    \end{bmatrix}
    \]
  </center>








## Matrix multiplication and Leslie matrix
  
  <center>
    \[
    \begin{bmatrix}
      n_{1,t+1} <br>
      n_{2,t+1} <br>
      n_{3,t+1} <br>
      n_{4,t+1}
    \end{bmatrix}
    =
    \begin{bmatrix}
      f_1 & f_2 & f_3 & f_4 <br>
      s_1 & 0 & 0 & 0 <br>
      0 & s_2 & 0 & 0 <br>
      0 & 0 & s_3 & 0
    \end{bmatrix}
    \times
    \begin{bmatrix}
      n_{1,t} <br>
      n_{2,t} <br>
      n_{3,t} <br>
      n_{4,t}
    \end{bmatrix}
    \]
  </center>





# Reproductive value




%% 
%%   ## Net reproductive rate $R_0$
%%   {\bf Definition}:  The average number of age class zero offspring
%%   produced by an average newborn organism during its lifetime
%%   
%%   \[
%%     R_0 =
%%   \]
%% 




  ## Reproductive Value
%  
%  {\bf Definition}: the number of future offspring expected
%  to be produced by an individual in age class $i$ over its remaining life
%  span, adjusted by the growth rate of the population \par
  {\bf Definition}: The extent to which an individual in age class
  $i$ will contribute to the ancestry of future generations. \par
%  
%  
%  \alert{Why adjust? \par}
  
%  
  \[
    v_i = \sum_{j=i}^{I}\left(\prod_{h=i}^{j-1}s_h\right)f_j\lambda^{i-j-1}
  \]
  
  
%  {\bf Fact}:
  Post-reproductive individuals have reproductive values of zero. <br>
  
  
  % {\bf Question}:
  Depending on survival and fecundity rates, a first-year individual
  could have a higher or lower reproductive value than a second-year
  individual. 







% 
%   ## Fisher's Reproductive Value
%   
%   {\bf Definition}: the number of future offspring expected
%   to be produced by an individual in age class $i$ over its remaining life
%   span, adjusted by the growth rate of the population \par
%   
%   
%   \alert{Why adjust? \par}
%   
%   
%   \[
%     v_i = \sum_{j=i}^{I}\left(\prod_{h=i}^{j-1}s_h\right)f_j\lambda^{i-j-1}
%   \]
% 




## Reproductive value for previous example

%  <center>
```{r}
#| label: rv
#| include: false
#| echo: false
V <- eigen(t(A))$vectors[,1]
rv <- as.numeric(V/V[1])
barplot(rv, names=c("Age class 1", "Age class 2", "Age class 3"),
        ylab="Reproductive value", ylim=c(0, 2), cex.names=1.5,
        cex.lab=1.5)
box()
```
%</center>

[width=0.8\textwidth]{figure/rv-1} <br>





% 
%   ## Reproductive value
%   Sir Ronald Aylmer Fisher was central to the development of the idea
%   of reproductive value.
%   <center>
%     [width=0.4\textwidth]{figs/fisher1} \hspace{12pt} %
%     [width=0.4\textwidth]{figs/fisher2}
%   </center>
%   He was also a terrible person (eugenicist and Nazi sympathizer).
% 




% 
%   ## Net reproductive rate ($R_0$)
%   Net reproductive rate is the expected number of offspring that will
%   be produced by an individual during its lifetime.
% \[
%   R_0 = \sum_{i=1}^I f_i \prod_{j=1}^{i-1} s_j
% \]
```{r}
#| echo: false
#| results: 'hide'
% f <- c(0, 0.5, 1.5, 1, 0.5, 0)
% s <- c(0.2, 0.3, 0.4, 0.4, 0.2, 0)
% R00 <- f[1]
% for(i in 2:length(f)) {
%     R00[i] <- f[i]*prod(s[1:(i-1)])
% }
% R0 <- sum(R00)
```
% 




% 
%   ## Generation time ($T$)
%   Generation time is the time required for a population to increase
%   by a factor of $R_0$.
% \[
%   T = \frac{\log(R_0)}{\log(\lambda)}
% \]

% 




## Other important quantities

**Net reproductive rate**

The expected number of individuals produced by an individual over  its lifetime:
$$
    R_0 = \sum_{i=1}^I \left(\prod_{j=1}^{i-1} s_j\right)f_i
$$
  
  
**Generation time**

The time required for a population to increase by a factor of
  $R_0$. %% Can also be defined as the age at which individuals are
  %% expected to reproduce.  
  % \[
  %   T = \frac{\log(R_0)}{\log(\lambda)}
  % \]




%
%  ## Concluding thoughts
%  
 % {\bf Future}
 % 
 %   * Sensitivity analysis: How does a change in $\theta$ affect $\lambda$?
 % 
 % 
%  {\bf Summary}
%  
%    * In some cases, size may be better descriptor of variation in
%      demographics
%    * We won't cover stage-based models
%    * Individual variation within age classes is ignored.
%  
%










  ## Other properties of the Leslie matrix\footnote{This is
      for graduate students only}
  
%  [<+->]
%  * The dominant eigenvalue of ${\bf A}$ is the growth rate $\lambda$. <br>
  
  
  % * The right eigenvector is the {\bf stable age distribution}. <br>
  
  
  % * The left eigenvector is the {\bf reproductive value}. <br>
%  






# Stage-structured models



## Stage-structured population models
  
%  
%  * Age isn't always the best way to think about population structure. <br>
  % * <2->
  
  
  For some populations, it is much more useful to think about size
      structure or even spatial structure. <br>
      % * <3->
  
  
  These ``stage-structured'' models differ from age-structured models
  in that individuals can remain in a stage class (with
  probability $1-p_i$) for multiple time periods. <br>
%  
%  
  \tikzstyle{level 1} = [circle, draw, text width=0.5cm, minimum size=1.5cm,
  node distance=3cm, text centered, fill=blue!10]
  <center>
   \uncover<4>{
   \begin{tikzpicture}[scale=0.8, every node/.style={scale=0.8}] \footnotesize %
      \node [level 1] (n1) [yshift=-5mm] {$n_{1,t}$};
      \node [level 1, right of=n1] (n2) [yshift=0mm] {$n_{2,t}$};
      \node [level 1, right of=n2] (n3) [yshift=0mm] {$n_{3,t}$};
      \draw[->,thick]  (n1) to node[above] {$s_1$} (n2);
      \draw[->,thick] (n2) to node[above] {$p_2s_2$} (n3);
      \draw[->,thick] (n2) to [bend right=40] node[above] {$f_2$} (n1);
      \draw[->,thick] (n3) to [bend right=60] node[above] {$f_3$} (n1);
      \draw[->,thick] (n2) to [loop below] node[below] {$(1-p_2)s_2$} (n2);
      \draw[->,thick] (n3) to [loop below] node[below] {$s_3$} (n3);
   \end{tikzpicture}
   }
  </center>







%\begin{comment}

  ## Stage-structured population models
  
  {In stage-structured models, individuals transition from one
    stage to the next with probability $p_i$. \par
    
    
    We can add these
    transition probabilities to our projection matrix like this:}
  
  %% 
  %% <center>
  %%   \[
  %%   {\bf A} =
  %%   \begin{bmatrix}
  %%     f_1 & f_2 & f_3 & f_4 <br>
  %%     s_1 & s_2(1-p_2) & 0 & 0 <br>
  %%     0 & s_2 p_2 & s_3(1-p_3) & 0 <br>
  %%     0 & 0 & s_3p_3 & s_4
  %%   \end{bmatrix}
  %%   \]
  %% </center>
  <center>
    
    \[
    {\bf A} =
    \begin{bmatrix}
      f_1 & f_2 & f_3 <br>
      s_1 & s_2(1-p_2) & 0 <br>
      0 & s_2 p_2 & s_3
    \end{bmatrix}
    \]
  </center>


This example allows for individual to remain in stages 2 and 3 for 
more than one time period.

%\end{comment}








  ## Summary
%  [<+->]
%  * Vital rates ($s$ and $f$) are usually age-specific. <br>
  
  
  % * Population growth will depend on age distribution. <br>
  
  
  % * If vital rates are constant, population will reach stable
  age distribution with constant growth rate $\lambda$. <br>
  
  
  % * Reproductive value indicates which age class contributes the
  most to population growth. <br>
  
  
  % * Matrix models are a convenient method used to work with
  age-structured populations. <br>
%  





\begin{comment}

# Sensitivity analysis



  ## Sensitivity analysis
  {\bf Questions}
  
    * How will $\lambda$ change if we change $b$ or $d$?
  






%\begin{comment}
# Life Tables



  ## Scenario
  
  
    * Age ($x$) begins at 0, ends at $k$
    * The age class ($i$) begins at 1, and ends at $s=k-1$
  
  <center>
    [width=0.6\textwidth]{figs/age-class-diagram}
  </center>
  
  
    * Individuals either advance to the next age class or die
    * An age class is not necessarily a 1 year time period
    * Each age class has its own vital rates
    * We usually only care about females
  





  ## Life Tables
  {\bf Assumptions}
  [(1)]
    * Birth-pulse model
      
        * Individuals give birth on day they enter new age class
      
    * Postbreeding census
  






  ## Life Tables
  <center>
    \begin{tabular}{cccccc}
      \hline
      Age ($x$) & Age class ($i$) & $b_x$ & $s_i$ & $l_x$         & $F_i$       <br>
      \hline
      0         &                 & 0     &          & 1.0          & <br>
      1         & 1               & 0.8   & 0.7      & $s_1$=0.7 & $b_1s_1$ <br>
      2         & 2               &       & 0.7      & $s_1s_2$=   & <br>
      \hline
    \end{tabular}
  </center>
  
    * [] $b_x$ is the average number of offspring produced by an
      individual of age $x$
    * [] $l_x$ is the probability of surviving from age 0 to age $x$
  







\note{Euler-Lotka equation: After stable age distribution is attained,
the population grows at a constant rate that is given by the solution
of the E-L equation. }







%% 
%%   ## Life tables in reverse
%%   \note{On the assumption that the population is (1) at a stable age
%%     distribution and (2) stationary (i.e. $\lambda$=1, it may be
%%     possible to use the standing age distribution to calculate
%%     age-specific survival rates}
%%   \note{In general, it is not possible to use age frequency data alone
%%     to both estimate age-specific survival and to test the assumptions
%%     of age stability and stationarity.}

%% 




\end{comment}





