---
format:
  revealjs:
    theme: default
    pagetitle: 'Logistic Population Growth'
    slide-number: true
    css: ../../custom.css
lang: en
---


# Background {visibility="hidden"}


## {content-align="center" style="font-size: 160%;" background-image="figs/turtles.jpg" background-opacity=0.2}

<br><br><br><br>

<center>
Logistic Growth

<br>

::: {style="font-size: 70%;"}
Applied Population Dynamics\
WILD 5700/7700
:::

</center>




## Today in APD

<br>

The equation for logistic growth in discrete time.

<br>
  
The definition of density-dependent growth.
  
<br>

Basic properties of the model.

<br>

Strange behavior of the (discrete time) model, such as damped oscillations and chaos.


# Logistic growth {visibility="hidden"}


## From Geometric to Logistic Growth

**Geometric growth** 

$$N_{t+1} = N_t + N_tr$$ 

. . .

**Logistic growth**

$$N_{t+1} = N_t + N_t r_{max}\left(1 - \frac{N_t}{K}\right)$$ 

. . .

where

::: {style="font-size: 80%;"}
- $r_{max}$ is the growth rate when $N_t$ is close to 0

- $K$ is the carrying capacity
:::


## Density-dependent growth

<br> 

Logistic growth is an example of **density-dependent growth**.

. . .

<br>
  
**Definition**: Population growth rate [is]{style="color: red;"} affected by population size ($N$).

. . .

<br>
  
**Implications**:  Resources are limited and there is a carrying capacity.




## Graphical depiction

```{r}
#| label: geomVlogistic
Time <- 0:100
T <- length(Time)
rmax <- 0.1
K <- 50
Ng <- Nl <- rep(0, T)
Ng[1] <- Nl[1] <- 2
for(t in 2:T) {
    Ng[t] <- Ng[t-1] + Ng[t-1]*rmax
    Nl[t] <- Nl[t-1] + Nl[t-1]*rmax*(1 - Nl[t-1]/K)
}
```

```{r}
#| label: geom
#| include: false
#| fig-height: 4.4
#| fig-width: 7
par(mai=c(0.9,0.9,0.1,0.1))
plot(Time, Ng, lwd=5, type="l", ylim=c(0, 100), cex.lab=1.5,
     xlab="Time (t)", ylab="Population size (N)")
legend(50, 100, c("Geometric growth", ""), lty=1, col=c("black", NA), lwd=4)
```

```{r}
#| label: geom-logistic
#| include: false
#| fig-height: 4.4
#| fig-width: 7
par(mai=c(0.9,0.9,0.1,0.1))
plot(Time, Ng, lwd=5, type="l", ylim=c(0, 100), cex.lab=1.5,
     xlab="Time (t)", ylab="Population size (N)")
lines(Time, Nl, lwd=4, col="purple")
legend(50, 100, c("Geometric growth", "Logistic growth"), lty=1,
       col=c("black","purple"), lwd=4)
```

<center>
::: {.r-stack}
::: {.fragment .fade-out fragment-index=1}
![](lecture-logistic-growth_files/figure-revealjs/geom-1)
:::
::: {.fragment .fade-in fragment-index=1}
![](lecture-logistic-growth_files/figure-revealjs/geom-logistic-1)
:::
:::
</center>



## Growth rate ($\lambda_t = N_{t+1}/N_t$)

```{r}
#| label: lr1
#| include: false
#| fig-width: 12
#| fig-height: 4
par(mai=c(0.9,1,0.3,0.3))
plot(Time, Nl, lwd=4, type="l", ylim=c(0, 60), cex.lab=1.9,
     xlab="Time (t)", ylab="Population size (N)", col="purple")
points(Time[which.min(abs(Nl-K/2))], Nl[which.min(abs(Nl-K/2))],
       pch=16, cex=5, col="orange")
text(Time[which.min(abs(Nl-K/2))]+5, Nl[which.min(abs(Nl-K/2))],
     labels="Inflection point (K/2)", pos=4, cex=1.5)
```

```{r} 
#| label: lambda
#| include: false
#| fig-width: 12
#| fig-height: 4
lambda <- Nl[-1]/Nl[-T]
par(mai=c(0.9,1,0.3,0.3))
plot(Time[-T], lambda, lwd=4, type="l", cex.lab=1.9,
     xlab="Time (t)", col="blue",
     ylab=expression(paste("Growth rate: ", "(", lambda, "=", N[]/N[], ")")))
```

<center>
<!-- ::: {.r-stack} -->
<!-- ::: {.fragment .fade-out fragment-index=1} -->
  ![](lecture-logistic-growth_files/figure-revealjs/lr1-1){width="80%"} 
<!-- ::: -->
<!-- ::: {.fragment .fade-in fragment-index=1} -->
  ![](lecture-logistic-growth_files/figure-revealjs/lambda-1){width="80%"}
<!-- ::: -->
</center>



## Growth ($\Delta_t=N_{t+1}-N_t$)

```{r}
#| label: lr2
#| include: false
#| fig-width: 12
#| fig-height: 4
par(mai=c(0.9,1,0.3,0.3))
plot(Time, Nl, lwd=4, type="l", ylim=c(0, 60), cex.lab=1.9,
     xlab="Time (t)", ylab="Population size (N)", col="purple")
points(Time[which.min(abs(Nl-K/2))], Nl[which.min(abs(Nl-K/2))],
       pch=16, cex=5, col="orange")
text(Time[which.min(abs(Nl-K/2))]+5, Nl[which.min(abs(Nl-K/2))],
     labels="Inflection point (K/2)", pos=4, cex=1.5)
```

```{r}
#| label: delta
#| include: false
#| fig-width: 12
#| fig-height: 4
delta <- Nl[-1] - Nl[-T]
par(mai=c(0.9,1,0.3,0.3))
plot(Time[-1], delta, lwd=4, type="l", cex.lab=1.9,
     xlab="Time (t)", col="turquoise4",
     ylab=expression(paste("Growth: ", "(", Delta, "=", N[]-N[], ")")))
```


<center>
<!-- ::: {.r-stack} -->
<!-- ::: {.fragment .fade-out fragment-index=1} -->
  ![](lecture-logistic-growth_files/figure-revealjs/lr2-1){width="80%"}
<!-- ::: -->
<!-- ::: {.fragment .fade-in fragment-index=1} -->
  ![](lecture-logistic-growth_files/figure-revealjs/delta-1){width="80%"}
<!-- ::: -->
</center>



## Growth ($\Delta_t=N_{t+1}-N_t$) as a function of $N$

```{r}
#| label: delta-N
#| include: true
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
plot(Nl[-T], delta, lwd=4, type="l", cex.lab=1.2,
     xlab="Population size (N)",
     ylab=expression(paste("Growth: ", "(", Delta, "=", N[]-N[], ")")))
abline(h=0, col="gray")
abline(v=K/2, lty=3, lwd=2)
text(K/2, 0.05, "Inflection point (K/2)", pos=4)
```

<!-- <center> -->
<!--   ![](lecture-logistic-growth_files/figure-revealjs/delta-N-1) -->
<!-- </center> -->










## What happens when we change $r_{max}$?

```{r}
Nl4 <- Nl3 <- Nl2 <- Nl
Nl4[1] <- Nl3[1] <- Nl2[1] <- 2
for(t in 2:T) {
    Nl2[t] <- Nl2[t-1] + Nl2[t-1]*0.5*(1-Nl2[t-1]/K)
    Nl3[t] <- Nl3[t-1] + Nl3[t-1]*2.0*(1-Nl3[t-1]/K)
    Nl4[t] <- Nl4[t-1] + Nl4[t-1]*3.0*(1-Nl4[t-1]/K)
}
```
```{r}
#| label: Nl
#| include: false
#| fig-width: 8
#| fig-height: 6
par(mai=c(0.9,0.9,0.5,0.1))
plot(Time, Nl, lwd=4, type="l", ylim=c(0, 100), cex.lab=1.3,
     xlab="Time (t)", ylab="Population size (N)", col="purple")
legend(0, 100, c("rmax=0.1", "", "", ""),
       lwd=4, col=c("purple", NA, NA, NA))
```
```{r}
#| label: Nl2
#| include: false
#| fig-width: 8
#| fig-height: 6
par(mai=c(0.9,0.9,0.5,0.1))
plot(Time, Nl, lwd=4, type="l", ylim=c(0, 100), cex.lab=1.3,
     xlab="Time (t)", ylab="Population size (N)", col="purple")
lines(Time, Nl2, lwd=4, col="blue")
legend(0, 100, c("rmax=0.1", "rmax=0.5", "", ""),
       lwd=4, col=c("purple", "blue", NA, NA))
```
```{r}
#| label: Nl3
#| include: false
#| fig-width: 8
#| fig-height: 6
par(mai=c(0.9,0.9,0.5,0.1))
plot(Time, Nl, lwd=4, type="l", ylim=c(0, 100), cex.lab=1.3,
     xlab="Time (t)", ylab="Population size (N)", col="purple",
     main="Damped oscillation")
lines(Time, Nl2, lwd=4, col="blue")
lines(Time, Nl3, lwd=4, col="orange")
legend(0, 100, c("rmax=0.1", "rmax=0.5", "rmax=2.0", ""),
       lwd=4, col=c("purple", "blue", "orange", NA))
```
```{r}
#| label: Nl4
#| include: false
#| fig-width: 8
#| fig-height: 6
par(mai=c(0.9,0.9,0.5,0.1))
plot(Time, Nl, lwd=4, type="l", ylim=c(0, 100), cex.lab=1.3,
     xlab="Time (t)", ylab="Population size (N)", col="purple",
     main="Chaos")
lines(Time, Nl2, lwd=4, col="blue")
lines(Time, Nl3, lwd=4, col="orange")
lines(Time, Nl4, lwd=4, col="gray")
legend(0, 100, c("rmax=0.1", "rmax=0.5", "rmax=2.0", "rmax=3.0"),
       lwd=4, col=c("purple", "blue", "orange", "gray"))
```

<center>
::: {.r-stack}
::: {.fragment .fade-out fragment-index=1}
![](lecture-logistic-growth_files/figure-revealjs/Nl-1){width="80%"}
:::
::: {.fragment .fade-in-and-out fragment-index=1}
![](lecture-logistic-growth_files/figure-revealjs/Nl2-1){width="80%"}
:::
::: {.fragment .fade-in-and-out fragment-index=2}
![](lecture-logistic-growth_files/figure-revealjs/Nl3-1){width="80%"}
:::
::: {.fragment .fade-in-and-out fragment-index=3}
![](lecture-logistic-growth_files/figure-revealjs/Nl4-1){width="80%"}
:::
:::
</center>






## Definitions 

::: {.beamer-block}

#### Overcompensation

::: {.beamer-block-content}
Density-dependent response in which populations over- or under-shoot carrying capacity rather than approach it gradually.
:::

:::

---

::: {.beamer-block}

#### Chaos

::: {.beamer-block-content}
Highly variable deterministic dynamics that are extremely sensitive to small changes in parameters.
:::

:::






## Assumptions of basic model
  
- $K$ and $r_{max}$ are constant

- No sex or age effects or other sources of individual heterogeneity

- No time lags

- No stochasticity







## Does logistic growth occur in nature?
  
  Fruit fly (*Drosophila melanogaster*) in the lab^[Data from Pearl (1927). Figure from Krebs (2014)]
  
<center>
![](figs/logistic-drosophila)
</center>







## Does logistic growth occur in nature?

Ibex (*Capra ibex*) in Switzerland^[Data from S&aelig;ther et al. (2002). Figure from Krebs (2014)]
  
<center>
[![](figs/logistic-ibex)](https://youtu.be/hqqCafpgSVA?si=greGPq4_wLm_BQLI)
</center>






# Summary {visibility="hidden"}




## Summary and assignment

**Summary**

::: {.incremental style="font-size: 75%;"}

- Logistic growth is a form of density-dependent growth

- Growth rate ($\lambda_t=N_{t+1}/N_t$) declines as $N$ approaches $K$

- Growth ($\Delta_t=N_{t+1}-N_t$) peaks at $K/2$ (the inflection point)

- The model doesn't include birth, mortality, and movement processes,
  but it can be derived from density-dependent birth and death rates.

- But it does allow for complex dynamics that resemble patterns seen in nature.
  
:::

. . . 

<br>

**Assignment**

<center>
Read pages 32--36 in Conroy and Carroll.
</center>



