---
format:
  revealjs:
    theme: default
    pagetitle: 'Logistic Population Growth'
    slide-number: true
    css: ../../custom.css
lang: en
---


# Background {visibility="hidden"}


## {content-align="center" style="font-size: 160%;" background-image="figs/turtles.jpg" background-opacity=0.2}

<br><br><br><br>

<center>
Logistic Growth

<br>

::: {style="font-size: 70%;"}
Applied Population Dynamics\
WILD 5700/7700
:::

</center>




## Learning objectives

The equation for logistic growth in discrete time.
  
The definition of density-dependent growth.
  
Basic properties of the model.
  
Strange behavior of the (discrete time) model, such as damped oscillations and chaos.


# Logistic growth {visibility="hidden"}


## From Geometric to Logistic Growth

**Geometric growth** 

$$N_{t+1} = N_t + N_tr$$ 

**Logistic growth**

$$N_{t+1} = N_t + N_t r_{max}\left(1 - \frac{N_t}{K}\right)$$ 

where

- $r_{max}$ is the growth rate when $N_t$ is close to 0

- $K$ is the carrying capacity



## Density-dependent growth

Logistic growth is an example of **density-dependent growth**
  
**Definition**: Population growth rate [is]  affected by population size ($N$).
  
**Implications**:  Resources are limited and there is a carrying capacity.




## Graphical depiction
```{r}
#| label: geomVlogistic
Time <- 0:100
T <- length(Time)
rmax <- 0.1
K <- 50
Ng <- Nl <- rep(0, T)
Ng[1] <- Nl[1] <- 2
for(t in 2:T) {
    Ng[t] <- Ng[t-1] + Ng[t-1]*rmax
    Nl[t] <- Nl[t-1] + Nl[t-1]*rmax*(1 - Nl[t-1]/K)
}
```

```{r}
#| label: geom
#| include: false
plot(Time, Ng, lwd=5, type="l", ylim=c(0, 100), cex.lab=1.5,
     xlab="Time (t)", ylab="Population size (N)")
legend(50, 100, c("Geometric growth", ""), lty=1, col=c("black", NA), lwd=4)
```

```{r}
#| label: geom-logistic
#| include: false
plot(Time, Ng, lwd=5, type="l", ylim=c(0, 100), cex.lab=1.5,
     xlab="Time (t)", ylab="Population size (N)")
lines(Time, Nl, lwd=4, col="purple")
legend(50, 100, c("Geometric growth", "Logistic growth"), lty=1,
       col=c("black","purple"), lwd=4)
```

<center>
::: {.r-stack}
::: {.fragment .fade-out fragment-index=1}
![](lecture-logistic-growth_files/figure-revealjs/geom-1)
:::
::: {.fragment .fade-in fragment-index=1}
![](lecture-logistic-growth_files/figure-revealjs/geom-logistic-1)
:::
:::
</center>




