\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0, 0, 0}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.69,0.494,0}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.749,0.012,0.012}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.514,0.506,0.514}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0,0.341,0.682}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.004,0.004,0.506}{#1}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}


%\usepackage[sans]{../pres1}

\usepackage[hmargin=1in,vmargin=1in]{geometry}
\usepackage{parskip}
%\hypersetup{pdfstartview=FitV,hidelinks}
\usepackage{verbatim}





%% knitr stuff


%% New command for inline code that isn't to be evaluated
\definecolor{inlinecolor}{rgb}{0.878, 0.918, 0.933}         % edit-kwrite
\newcommand{\inr}[1]{\colorbox{inlinecolor}{\texttt{#1}}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

{
  \Large
  \centering
  {\bf Lab 9 Assignment --- Distance Sampling} \\
  Due before Monday \par
}

\vspace{10pt}

% Analysis of the (fake) Mongolian gazelle ({\it Procapra gutturosa})
% data using program DISTANCE.  

The purpose of this lab is to learn how to use the
R package `Distance' to estimate abundance and density. You will
do this by analyzing fake data on Mongolian gazelle ({\it Procapra
  gutturosa}). The data are fake, but they are based loosely on the
study that we discussed in lecture. Note that gazelles were detected
in groups (ie, herds), and the distance data are distances to group
centers. 

Put your answers in a Word file and upload it to ELC. Name the file
something like ``Chandler-lab9.docx''. Due before your next lab.  





\section*{R package `Distance'}

Open R (or RStudio) and install the package using the following
command: 

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.878, 0.918, 0.933}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{install.packages}\hlstd{(}\hlstr{"Distance"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

Several other packages will be installed during the process. Now we
can load the package like so: 

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.878, 0.918, 0.933}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{library}\hlstd{(Distance)}
\end{alltt}
\end{kframe}
\end{knitrout}

Before we can do any analysis, we need to import the Mongolian gazelle
data. Make sure that the file \texttt{GazelleFakeData.txt} is in your
working directory.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.878, 0.918, 0.933}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{getwd}\hlstd{()}       \hlcom{## Location on your computer where R will look for files}
\end{alltt}
\begin{verbatim}
## [1] "/home/rchandler/courses/applied-popdy/labs/distance"
\end{verbatim}
\begin{alltt}
\hlkwd{list.files}\hlstd{()}  \hlcom{## You should see 'GazelleFakeData.txt' here}
\end{alltt}
\begin{verbatim}
##  [1] "auto"                "figs"                "GazelleFakeData.txt"
##  [4] "lab-distance-R.aux"  "lab-distance-R.log"  "lab-distance-R.pdf" 
##  [7] "lab-distance-R.Rnw"  "lab-distance-R.tex"  "lab-distance.pdf"   
## [10] "lab-distance.Rnw"    "lab-distance.tex"
\end{verbatim}
\end{kframe}
\end{knitrout}

If you need to change your working directory, you can use the dropdown
menu options, or you can use a command like this:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.878, 0.918, 0.933}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{setwd}\hlstd{(}\hlstr{"C:/Users/RichardC/courses/"}\hlstd{)} \hlcom{## Change the path in quotes!}
\end{alltt}
\end{kframe}
\end{knitrout}

Once you have the data file in your working directory, you can import
the fake gazelle data. The data file is a tab-separated text file
instead of a comma-seperated text file like we used last week:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.878, 0.918, 0.933}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{gazelleData} \hlkwb{<-} \hlkwd{read.table}\hlstd{(}\hlstr{"GazelleFakeData.txt"}\hlstd{,} \hlkwc{header}\hlstd{=}\hlnum{TRUE}\hlstd{,} \hlkwc{sep}\hlstd{=}\hlstr{"\textbackslash{}t"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}


\newpage

The structure of the data can be assessed like this:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.878, 0.918, 0.933}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{str}\hlstd{(gazelleData)}
\end{alltt}
\begin{verbatim}
## 'data.frame':	118 obs. of  6 variables:
##  $ Transect      : int  1 1 1 1 1 1 1 1 1 2 ...
##  $ GroupID       : int  1 2 3 4 5 6 7 8 9 1 ...
##  $ Distance      : int  23 23 17 40 87 77 50 62 3 33 ...
##  $ GroupSize     : int  2977 2908 2899 2994 2998 3087 2992 3057 2928 3030 ...
##  $ TransectLength: int  100 100 100 100 100 100 100 100 100 100 ...
##  $ RegionArea    : int  5 5 5 5 5 5 5 5 5 5 ...
\end{verbatim}
\end{kframe}
\end{knitrout}

We have to reformat the data to meet the requirements of the Distance
package. The code below creates a new data frame, renames the columns,
and converts the distance units to kilometers.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.878, 0.918, 0.933}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{gazelleData2} \hlkwb{<-} \hlstd{gazelleData[,}\hlkwd{c}\hlstd{(}\hlstr{"Transect"}\hlstd{,} \hlstr{"GroupSize"}\hlstd{)]}
\hlkwd{colnames}\hlstd{(gazelleData2)} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"Sample.Label"}\hlstd{,} \hlstr{"size"}\hlstd{)}
\hlstd{gazelleData2}\hlopt{$}\hlstd{distance} \hlkwb{<-} \hlstd{gazelleData}\hlopt{$}\hlstd{Distance} \hlopt{/} \hlnum{1000} \hlcom{## Convert to km}
\hlstd{gazelleData2}\hlopt{$}\hlstd{Region.Label} \hlkwb{<-} \hlnum{1}
\hlstd{gazelleData2}\hlopt{$}\hlstd{Effort} \hlkwb{<-} \hlstd{gazelleData}\hlopt{$}\hlstd{TransectLength}
\hlstd{gazelleData2}\hlopt{$}\hlstd{Area} \hlkwb{<-} \hlstd{gazelleData}\hlopt{$}\hlstd{RegionArea}
\end{alltt}
\end{kframe}
\end{knitrout}


The distance sampling model can be fitted to the data using the
\inr{ds} function. The \inr{key} argument specifies the detection
function. You could switch from the half-normal to the hazard-rate
function using: \inr{key="hr"}. The \inr{truncation} argument
indicates that we didn't record data beyond 250 m (0.25 km).

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.878, 0.918, 0.933}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{model.hn} \hlkwb{<-} \hlkwd{ds}\hlstd{(}\hlkwc{data}\hlstd{=gazelleData2,} \hlkwc{key}\hlstd{=}\hlstr{"hn"}\hlstd{,}
               \hlkwc{transect}\hlstd{=}\hlstr{"line"}\hlstd{,} \hlkwc{truncation}\hlstd{=}\hlnum{250}\hlopt{/}\hlnum{1000}\hlstd{,}
               \hlkwc{adjustment}\hlstd{=}\hlkwa{NULL}\hlstd{,} \hlkwc{quiet}\hlstd{=}\hlnum{TRUE}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}


Once you have fitted the model, you can obtain the parameter estimates
using the following command:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.878, 0.918, 0.933}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{summary}\hlstd{(model.hn)}
\end{alltt}
\end{kframe}
\end{knitrout}

The estimate of the scale parameter $\sigma$ and average detection
probability $\bar{p}$ can be found near the top of the output under
\verb+Detection function parameters+. The scale parameter is estimated
on the natural log scale, but it can be transformed to the original
scale with the exponential function, for example using code like this:
\inr{exp(-2.549)}. 

The estimates of abundance and density are toward the bottom of the
output under \\
\verb+Summary for individuals+.
 

To view the estimated detection function, do this:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.878, 0.918, 0.933}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{plot}\hlstd{(model.hn)}
\end{alltt}
\end{kframe}
\end{knitrout}


\clearpage

\section*{Assignment -- Compare results from half-normal and hazard-rate models}

\begin{enumerate}
  \item Run the model above with the half-normal detection function.
  \item Run another model, but use a hazard-rate detection function
    instead.  
  \item Create a table in Word to compare the two models. The table
    should have a row for each model. The column names should be:
    \texttt{Model, pbar, Abundance, SE(N), Density, SE(D), AIC}.
    The two \verb+SE+ columns are for the standard errors of
    abundance and density. 
  \item Create a figure with the two graphs of the detection functions
    side-by-side.
  \item Write a short paragraph summarizing your results. Indicate
    which model was better according to AIC, and explain why 
\end{enumerate}





\end{document}




